Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-07-12 write-up/covid-writeup.tex
Update covid-writeup.tex
Edits through section 3.4

\usepackage[foot]{amsaddr}
% \usepackage{biblatex}
\usepackage{natbib}
%\bibliographystyle{apalike}
\usepackage{multirow}
\usepackage{enumitem}

@@

Due to monetary and time constraints, however, random testing is either not performed or performed very infrequently.  As of June 2021, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results. The CDC has conducted several seroprevalence studies. See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

@@

Note that the proposed approach requires covariate information to be collected on tested individuals.  Unfortunately, many states do not require or report auxiliary covariate information beyond demographics such as gender, age, race, and ethnicity.  To address this, we discuss how the proposal suggests minor changes to current practice that may improve performance over time. While we demonstrate empirical improvements over simple disease prevalence estimates, we also highlight how selection bias may persist and impact uncertainty quantification.

@@

\subsection{Outline}

In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.  We highlight how testing information is collected as well as demographic and comorbidity information.  In Section~\ref{section:testinginfo}, we briefly summarize COVID-19 testing.  We clarify the difference between viral and antibody testing.  For viral testing, we briefly discuss the reasons behind measurement-error in RT-PCR tests including test timing and sample collection techniques.

In Section~\ref{section:casecount}, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.

In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.  Weight estimation requires auxiliary information provided by random samples.  A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in effective data quantity.

In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 2020 until June 2021.  We estimate data quality for IPW estimates using the random sample on April 28th.  This demonstrates the proposed approach is an improvement over simple prevalence estimation but that data quality is not of the same order as random sampling.

In Section~\ref{section:discussion}, a series of recommendations are provided that aim to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.

@@

In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, their respective scientific uses in managing the pandemic, and the data streams we consider in this manuscript.

@@

A molecular test conducted on an infected individual may return a negative result.  Such false negatives are very strongly associated with when the test is conducted.  In the incubation phase, most molecular tests will return a false negative result.  Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility~\citep{Mina2020}.   Moreover, most molecular tests are performed via nasopharyngeal swab.  Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.   Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020}.

@@

Our primary goal is accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.  Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.  A final goal is characterizing prevalence of population-level immunity. These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
 % as well as the impact of public health  interventions.

@@

Coronavirus case-count data refers to the number of positive molecular tests performed on each day.  Figure~XX plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~XX, dotted blue line). On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.

Public databases most often contain aggregate information.  The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.  Aggregate time series of case count and deaths can also be extracted.  Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time. Working closely with the State of Indiana, we were able to generate publicly available datasets for COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.  This granular dataset motivates the proposed approach.
% https://www.regenstrief.org/covid-dashboard/
%

@@

Due to testing restrictions and other potential selection biases, publicly reported covid testing data may not be sufficient to understand active infection rates or the disease trajectory.  Here, we discuss two random samples that can be leveraged to get either unbiased estimates of the statewide active infection rate or as auxiliary information to address selection bias in the testing data.

@@

\subsubsection{Facebook/CMU symptom survey.}
Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.  Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.  The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.  Figure~\ref{fig:fbsymptoms} displays nonparametric regression estimates of the fraction of individuals within age and gender strata.  In this paper, we use this as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of popopulation displaying symptoms or in contact with COVID-19 positive individuals).

\begin{figure}[!th]
\centering
\includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
\caption{Reported rate of COVID-like symptoms in strata.  Daily rates were estimated using weighted method on each day separately and then smoothed over time using local-linear nonparametric regression.}
\label{fig:fbsymptoms}
\vspace{-0.3cm}
\end{figure}

@@

The above selection mechanisms are random and independent of the outcome of interest. When this is not the case, selection effects may cause bias in the above estimates. To better understand this issue, \cite{Meng2018} provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
$$
The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}. The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.

@@

Tests are imperfect.  As discussed in Section~\ref{section:testinginfo}, COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  In discussions of inaccurate testing, the standard assumption is measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.  In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity were known a priori, we present a novel iterative procedure that uses the error decomposition to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators. To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
\begin{equation}
\label{eq:statdecomp}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
\end{equation}
where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.  See Appendix~\ref{app:memestimator} for the derivation. This extends work by \cite{Meng2018} to account for imperfect testing. The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.  The final term represents the \emph{imperfect testing adjustment} which is again a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$ compared to prior decomposition.

Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}. Under perfect testing, $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates. When tests are imperfect, this simple relationship no longer holds.

\begin{figure}[!th]
\centering
\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$. \textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{figure}

@@

The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.  Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time.  If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.  In this section, we demonstrate how selection bias and imperfect testing impact such estimates.

Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.  Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/Y_j$ is the coefficient of variation on day $j$.  See Appendix~\ref{app:ratio} for the derivation. The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a relatively small error compared to a large increase. The second term represents potential \emph{cancellation} which may occur when data quality, sampling fraction, measurement error, and prevalence are all constant across time.

Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
\begin{equation}
\label{eq:sir}
\frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
\frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
\frac{\partial r_t}{\partial t} = - \gamma i_t
\end{equation}
where $s_t$ is the fraction of susceptible individuals, $i_t$ is the fraction of infected individuals, and $r_t$ is the fraction of removed (recovered or deceased) individuals in the population at time $t$.  In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.  From a decision making perspective, such biases have a substantial impact.  First, the overestimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data may be pessimistic. What is missing in the current public health discourse is that the bias direction of the prevalence estimate is non-constant over time.   Underestimation post-peak puts pressure on lawmakers and governors to prematurely relax social distancing measures.  Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.  Note, however, that the standard errors will be large due to small effective sample sizes.

@@

\begin{lemma}[Variance of IPW estimator] \normalfont
\label{lemma:ipw}
\textcolor{red}{Must check.}
The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

Let $s_t$, $e_t$, $i_t^{(T)}$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.  The population-level SEIR dynamics are then given by the following set of differential equations:
\begin{align*}
\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \\
\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
\frac{\partial r_t}{\partial t} = - \gamma i_t.
\end{align*}
Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization. Here, we consider covariate information to take the form of a stratificatoin variable $k$ with $K$ strata.  Let $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.  AS there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution
$$
( \nu_{1,t}, \ldots, \nu_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{i1,t}, \ldots, p_{iK, t}) \right).
$$

>>>>

\usepackage[foot]{amsaddr}
% \usepackage{biblatex}
\usepackage{natbib}
%\bibliographystyle{apalike}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{caption,subcaption}

@@

Due to monetary and time constraints, however, random testing is either not performed or performed infrequently.  As of June 2021, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results. The CDC has conducted several seroprevalence studies. See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

@@

The proposed approach requires covariate information to be collected on individuals who receive a COVID-19 test.  Unfortunately, many states do not require or report auxiliary covariate information beyond basic demographic information (e.g., gender, age, race, and ethnicity).  We end with a brief list of suggestions of minor changes to current practice based on the proposed methodology. While we demonstrate empirical improvements over simple disease prevalence estimates, we also highlight how selection bias may persist and impact uncertainty quantification.

@@

% \subsection{Outline}

% In Section~\ref{section:data}, we summarize the publicly available data on COVID-19.  We highlight how testing information is collected as well as demographic and comorbidity information.  In Section~\ref{section:testinginfo}, we briefly summarize COVID-19 testing.  We clarify the difference between viral and antibody testing.  For viral testing, we briefly discuss the reasons behind measurement-error in RT-PCR tests including test timing and sample collection techniques.

% In Section~\ref{section:casecount}, we present simple mathematical arguments to show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.

% In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.  Weight estimation requires auxiliary information provided by random samples.  A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in effective data quantity.

% In Section~\ref{section:applications}, the IPW and doubly-robust estimation methods are applied to the case study of Indiana statewide prevalence from March 2020 until June 2021.  We estimate data quality for IPW estimates using the random sample on April 28th.  This demonstrates the proposed approach is an improvement over simple prevalence estimation but that data quality is not of the same order as random sampling.

% In Section~\ref{section:discussion}, a series of recommendations are provided that aim to help improve statistical inference in settings with large nonprobability samples from a stochastic process where small random sampling over time is possible.

@@

In this section we provide the necessary background to understand COVID-19 diagnostic and serologic testing, discuss their respective scientific uses in managing the pandemic, and introduce the data streams considered in this manuscript.

@@

A molecular test conducted on an infected individual may return a negative result.  Such false negatives are very strongly associated with when the test is conducted.  In the incubation phase, most molecular tests will return a false negative result.  Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility~\citep{Mina2020}.   Moreover, most molecular tests are performed via nasopharyngeal swab.  Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.   Systematic reviews suggest that 70\% sensitivity and 97.6\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020,Cohen2020}.

@@

The primary goal of this manuscript is to produce accurate estimates of the population-level active infection rates over time using \emph{publicly available} viral testing data.  Secondary goals include estimation of rates of change and the effective reproduction number which characterize disease trajectory.  A final goal is characterizing prevalence of population-level immunity. These quantities are fundamental to public health policy and provide critical information on the presence and transmission of SARS-CoV-2.
 % as well as the impact of public health  interventions.

@@

Coronavirus case-count data refers to the number of positive molecular tests performed on each day.  Figure~\ref{fig:in-cases} plots the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Figure~\ref{fig:in-tests} plots the total number of COVID-19 molecularw tests performed per day. Figure~\ref{fig:in-deaths} plots the total number of COVID-19 related reported deaths per day. Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.

Public databases most often contain aggregate information.  The Johns Hopkins dashboard, for example, provides demographic breakdown of case counts as well as the total confirmed cases and deaths by county.  Aggregate time series of case count and deaths can also be extracted.  Unfortunately, most dashboards do not provide any demographic information on who requested a test nor on who tested positive for SARS-CoV-2 over time. Working closely with the State of Indiana, we were able to produce publicly available datasets for COVID-19 total tests, positive tests, and related deaths per day broken out marginally by age, gender, ethnicity, and race~\citep{IndianaData2021}.  This granular dataset motivates the proposed approach.
% https://www.regenstrief.org/covid-dashboard/
%

@@

Due to testing restrictions and other potential selection biases, publicly reported Covid testing data may not be sufficient to understand active infection rates or the disease trajectory.  Here, we discuss two random samples that can be leveraged to as auxiliary information to address selection bias.

@@

\subsubsection{Facebook/CMU symptom survey.}
Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.  Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.  The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.  Figure~\ref{fig:fbsymptoms} displays smoothed estimates of the fraction of individuals within age and gender strata who report a fever within the past 24-hours.  This paper uses the Facebook symptom survey as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of population displaying symptoms or in contact with COVID-19 positive individuals).

\begin{figure}[!th]
\centering
\includegraphics[width = 0.9\textwidth]{../figs/fbcovid19symptoms.png}
\caption{Rate of reported fever per strata.  Daily rates were estimated using weighted method on each day separately and then smoothed over time using local-linear nonparametric regression.}
\label{fig:fbsymptoms}
\vspace{-0.3cm}
\end{figure}

%% GOOD THROUGH HERE

@@

These random selection mechanisms are independent of the outcome of interest. When this is not the case, selection effects may cause bias. To better understand this issue, \cite{Meng2018} recently provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
$$
The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}. The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.

@@

Tests are imperfect.  As discussed in Section~\ref{section:testinginfo}, COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  Researchers most often assume measurement error will lead to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we incorrectly measure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^{\star} = Y_j(1-P_j) + (1-Y_j) P_j$.  The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples.  In Appendix~\ref{app:memestimator}, assuming sensitivity and specificity were known a priori, we present a novel iterative procedure that uses the error decomposition to construct the estimator $\tilde y_n = (\bar y_n - FP)/(1-(FP+FN))$, which is unbiased under simple random sampling (SRS); see Appendix~\ref{app:modelbased} for a discussion of the connection to model-based estimators. To understand the impact of selection bias and imperfect testing, we derive the following statistical decomposition of the error between $\tilde y_n$ and $\bar Y$:
\begin{equation}
\label{eq:statdecomp}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M},
\end{equation}
where $f=n/N$ is the sampling fraction, $f_1$ and $f_0$ are sampling fractions for COVID-19 positive and negative individuals respectively, and $\Delta = f_1 - f_0$ is the sampling rate differential.  See Appendix~\ref{app:memestimator} for the derivation. This extends work by \cite{Meng2018} to account for imperfect testing. The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}.  The final term represents the \emph{imperfect testing adjustment} which is a complex function of the sampling rate differential, the odds ratio, and the ratio of measurement error interaction with prevalence and sampling rates interaction with prevalence but here is scaled by the factor $(1 - FP - FN)^{-1}$.

Figure~\ref{fig:heatmap} shows that $D_M$, as a function of the relative frequency ($f_1/f_0$) and odds ratio, can be both positive and negative as well as a range of magnitudes~\citep{Beesley2020,Beesley2019,Smeden2019}. Under either perfect testing (i.e., $FP=FN=0$), $D_M = 1$ so the relation between estimation and selection bias is simple, e.g., COVID-19 positive individuals more likely to receive test implies upward bias in prevalence estimates. Under random testing (i.e., $f_0 = f_1$), $D_M = (1-FP-FN)^{-1}$ so measurement error simply scales the error. When tests are imperfect and selection bias exists, this simple relationship no longer holds.

\begin{figure}[!th]
\centering
\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.024$ and $FN=0.30$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
%\textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{figure}

@@

The prior analysis demonstrates the limited information regarding COVID-19 prevalence in observational case-count data.  Scientists, however, may claim that daily observed case-counts simply undercount daily total cases by a constant multiple over time.  If true then the ratio of case-counts at consecutive times may be a good estimate of the true change in prevalence, helping scientists understand the disease trajectory.  We next demonstrate how selection bias and imperfect testing impact such estimates.

Let $\bar Y_{t-1}$ and $\bar Y_{t}$ denote the prevalence on two consecutive days and consider the estimator $r = \tilde y_t / \tilde y_{t-1}$.  Using a second-order Taylor series approximation, the error between ${\tilde y_t}/{\tilde y_{t-1}}$ and ${\bar Y_{t}}/{\bar Y_{t-1}}$ can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_t}{\bar Y_{t-1}} &\times \bigg[ \rho_{I_t,Y_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (Y_t)  -\rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},Y_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (Y_{t-1}) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement error adjustment, and $CV(Y_j) = \sigma_{Y_j}/ \bar Y_j$ is the coefficient of variation on day $j$.  See Appendix~\ref{app:ratio} for the derivation. The error magnitude depends on the true rate $\bar Y_{t} / \bar Y_{t-1}$ so a large decrease will have a small error relative to a large increase. The second term represents potential \emph{cancellation} which can  occur when data quality, sampling fraction, measurement error, and prevalence are constant across time.

Figure~\ref{fig:ratio-bias} displays the trajectory of the true ratio and the potential biased estimators under an SIR model~\citep{Pastor2001,Newman2002,Parshani2010} for the epidemic dynamics, with state evolution given by
\begin{equation}
\label{eq:sir}
\frac{\partial s_t}{\partial t} = - \beta s_t i_t; \quad
\frac{\partial i_t}{\partial t} = \beta s_t i_t - \gamma i_t; \quad
\frac{\partial r_t}{\partial t} = - \gamma i_t
\end{equation}
where $s_t$ is the fraction of susceptible individuals, $i_t$ is the fraction of infected individuals, and $r_t$ is the fraction of removed (recovered or deceased) individuals in the population at time $t$.  In terms of bias, the rate is overestimated prior to the peak and underestimated afterwards; the bias increases dramatically when the relative fraction exceeds $2$.  Such biases may have a substantial impact on policy making.  First, overestimation may give policy makers more leverage in proposing aggressive actions to reduce prevalence.  Of course, the analysis provides evidence that estimates based on available data may be pessimistic. What is missing in the current discourse is that the direction of bias is non-constant over time.   Underestimation post-peak puts pressure on policy makers to prematurely relax social distancing measures.  Moreover, it appears that the peak time is the easiest to estimate in terms of having minimal error using available data.  Note, however, that the standard errors will be large due to small effective sample sizes.

%%  MADE IT TO HERE

@@

\begin{lemma}[Variance of IPW estimator] \normalfont
\label{lemma:ipw}
% \textcolor{red}{Must check.}
The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

Let $s_t$, $e_t$, $i_t^{(T)}$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.  The population-level SEIR dynamics are then given by the following set of differential equations:
\begin{align*}
\frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \\
\frac{\partial i^{(T)}_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
\frac{\partial r_t}{\partial t} = - \gamma i_t.
\end{align*}
Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization. Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.  Let $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.  As there are a total of $N \cdot i_t := I_t$ new infections on day $t$, we assume the joint distribution of strata-specific new infection totals takes a multinomial distribution
$$
( \nu_{1,t}, \ldots, \nu_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
$$
